Comparing the Direction and Degree of Selection on Mutations Between SARS-CoV-2 Phylogenetic Clades

### 2022-10-26 notebooks/expected_and_actual_counts_ratio.Rmd
Modify code and restructure data frames to speed up time-consuming commands

# Define a minimum threshold for the number of expected counts. Mutations below this threshold will be removed.
min_expected <- 10

@@

```{r temp}
filtered_comparisons$clade1_observed <- fisher_data[which(fisher_data$nt_mutation_clade %in% filtered_comparisons$clade1),"observed"]
filtered_comparisons$clade1_expected <- fisher_data[which(fisher_data$nt_mutation_clade %in% filtered_comparisons$clade1),"expected"]
filtered_comparisons$clade2_observed <- fisher_data[which(fisher_data$nt_mutation_clade %in% filtered_comparisons$clade2),"observed"]
filtered_comparisons$clade2_expected <- fisher_data[which(fisher_data$nt_mutation_clade %in% filtered_comparisons$clade2),"expected"]

# alternate p-value calculations between sequential rows
p_values_alt <- data.frame(p_value=apply(filtered_comparisons, 1, FUN=function(c) {fisher.test(data.frame(clade1=c(as.numeric(c[3]),as.numeric(c[4])),clade2=c(as.numeric(c[5]),as.numeric(c[6]))))$p.value}))
```

@@

```{r contingency_analysis}
# Save list of p-values for every comparison. NOTE: Expected counts are not integers and are therefore rounded for the Fisher's Exact test. Mac Desktop Local: "334.273  41.641 375.678"
p_values <- data.frame(p_value=apply(filtered_comparisons, 1, FUN=function(c) {fisher.test(fisher_data[c(which(fisher_data$nt_mutation_clade==c[1]), which(fisher_data$nt_mutation_clade==c[2])), c("observed","expected")])$p.value}))

# Save and load variable because calculating p-values takes a long time.
'save(p_values, file="fisher_pvalues.Rda")'
load("fisher_pvalues.Rda")

# Combine p-value list with the clade comparison list.
comparison_p_values <- cbind(filtered_comparisons, p_values)
```

@@

```{r p_value_correction}
# False Discovery Rate and Bonferroni Corrections

@@

```{r fold-change}
# Calculate fold change for plotting volcano plots [clade 1 observed/clade 1 expected]/[clade 2 observed/clade 2 expected].
comparison_p_values$fold_change <- apply(comparison_p_values[,c("clade1","clade2")], 1, FUN=function(c) {foldchange(fisher_data[which(fisher_data$nt_mutation_clade==c[1]), "observed_expected"], fisher_data[which(fisher_data$nt_mutation_clade==c[2]), "observed_expected"])})
```

>>>>

# Define a minimum threshold for the number of expected counts. Mutations below this threshold will be removed.
min_expected <- 0

@@

To decrease the amount of time it takes to run the contingency tests, add columns to the filtered comparisons data frame with the observed and expected values for clade 1 and clade 2.

```{r contingency_matrix}
# Add observed and expected values for each comparison to filtered comparisons data frame.
{filtered_comparisons$clade1_observed <- fisher_data[which(fisher_data$nt_mutation_clade %in% filtered_comparisons$clade1),"observed"]
filtered_comparisons$clade1_expected <- fisher_data[which(fisher_data$nt_mutation_clade %in% filtered_comparisons$clade1),"expected"]
filtered_comparisons$clade2_observed <- fisher_data[which(fisher_data$nt_mutation_clade %in% filtered_comparisons$clade2),"observed"]
filtered_comparisons$clade2_expected <- fisher_data[which(fisher_data$nt_mutation_clade %in% filtered_comparisons$clade2),"expected"]}

#
{filtered_comparisons$clade1_obs_exp <- fisher_data[which(fisher_data$nt_mutation_clade %in% filtered_comparisons$clade1),"observed_expected"]
filtered_comparisons$clade2_obs_exp <- fisher_data[which(fisher_data$nt_mutation_clade %in% filtered_comparisons$clade2),"observed_expected"]}
```

@@

```{r contingency_table}
# Iterate through rows and compare clade 1 observed and expected values to clade 2 observed and expected values.
# NOTE: Expected counts are not integers and are therefore rounded for the Fisher's Exact test. 
'p_values <- data.frame(p_value=apply(filtered_comparisons, 1, FUN=function(c) {fisher.test(rbind(c(as.numeric(c[3]),as.numeric(c[4])),c(as.numeric(c[5]),as.numeric(c[6]))))$p.value}))'

# Save and load variable because calculating p-values takes a long time.
'save(p_values, file="fisher_pvalues.Rda")'
load("fisher_pvalues.Rda")
```

Combine the p-values calculated using the Fisher's Exact test with the data frame of filtered comparisons.

```{r combine_p_values}
# Combine p-value list with the clade comparison list.
comparison_p_values <- cbind(filtered_comparisons, p_values)
```

@@

```{r p_value_correction}
# False Discovery Rate and Bonferroni Corrections
# NEED TO CORRECT FOR MORE THAN 2 CLADE COMPARISONS
comparison_p_values$p_values_corrected <- comparison_p_values$p_value/nrow(comparison_p_values)
```

@@

```{r fold-change}
# Calculate fold change for plotting volcano plots [clade 1 observed/clade 1 expected]/[clade 2 observed/clade 2 expected].
comparison_p_values$fold_change <- foldchange(comparison_p_values$clade1_obs_exp, comparison_p_values$clade2_obs_exp)
```
